---
title: "Cumulative Incidence Curve Regression"
author: "Zachary McCaw"
date: "`r Sys.Date()`"
output:
  pdf_document
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Pseudo-value regression}
  %\usepackage[UTF-8]{inputenc}
---

## Purpose

This vignette illustrates regression of cumulative incidence probabilities via pseudo-values, using **CICs** for the CIC and pseudo-value calculations.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
suppressPackageStartupMessages({
  library(CICs)
  library(dplyr)
  library(ggplot2)
})
```

## Pseudo-value regression

### Simulate data

We generate two-arm competing risks data using **CICs** `GenTwoSampleData`, which calls `GenData` for each arm. Arm 1 has a lower event rate (0.35) than arm 0 (0.50), with the same death and censoring rates in both arms.

```{r}
set.seed(101)
df <- GenTwoSampleData(
  n1 = 200,
  n0 = 200,
  event_rate1 = 0.35,
  event_rate0 = 0.50,
  death_rate1 = 0.25,
  death_rate0 = 0.25,
  censor_rate1 = 0.25,
  censor_rate0 = 0.25,
  tau = 10
)

df %>%
  dplyr::group_by(arm) %>%
  dplyr::summarise(
    n = dplyr::n(),
    n_censor = sum(status == 0),
    n_event = sum(status == 1),
    n_death = sum(status == 2),
    .groups = "drop"
  )
```

### Pseudo-values

The pseudo-value for subject $i$ at time $\tau$ is
$$\hat{\theta}_i(\tau) = n \cdot \hat{\theta}(\tau) - (n-1) \cdot \hat{\theta}_{(-i)}(\tau),$$
where $\hat{\theta}(\tau)$ is the CIC at $\tau$ from the full sample and $\hat{\theta}_{(-i)}(\tau)$ is the CIC at $\tau$ from the sample with subject $i$ removed. We use **CICs** `CalcCIC` for both.

```{r}
# CIC at time tau (step function: value at last time <= tau).
cic_at_tau <- function(time, status, tau) {
  fit <- CalcCIC(status = status, time = time)
  idx <- which(fit$time <= tau)
  if (length(idx) == 0) return(0)
  fit$cic_event[max(idx)]
}

# Pseudo-values at tau via jackknife (using CICs::CalcCIC).
gen_pseudo_cic <- function(data, tau) {
  n <- nrow(data)
  time <- data$time
  status <- data$status
  theta_full <- cic_at_tau(time, status, tau)
  theta_loo <- vapply(seq_len(n), function(i) {
    cic_at_tau(time[-i], status[-i], tau)
  }, numeric(1))
  data$pseudo <- n * theta_full - (n - 1) * theta_loo
  data
}
```

### Fit pseudo-value regression

Pseudo-values at $\tau = 2$ are modeled with linear regression on treatment arm; the coefficient is the risk difference at $\tau$.

```{r}
tau <- 2
df <- gen_pseudo_cic(df, tau = tau)
fit <- stats::lm(pseudo ~ arm, data = df)
summary(fit)
```

The negative coefficient on arm indicates that treatment (arm 1) is associated with a lower cumulative incidence of the event at $\tau = 2$, consistent with the lower event rate in arm 1.

### Ground truth (no censoring)

With no censoring, a larger sample gives a precise estimate of the risk difference at $\tau$. We use `GenTwoSampleData` with `censor_rate` 0 and a truncation time so that all events are observed.

```{r}
set.seed(101)
df_gt <- GenTwoSampleData(
  n1 = 500,
  n0 = 500,
  event_rate1 = 0.35,
  event_rate0 = 0.50,
  death_rate1 = 0.25,
  death_rate0 = 0.25,
  censor_rate1 = 0,
  censor_rate0 = 0,
  tau = 10
)
df_gt <- gen_pseudo_cic(df_gt, tau = 2)
fit_gt <- stats::lm(pseudo ~ arm, data = df_gt)
summary(fit_gt)
```
